Concurrent Programming with CXL

### 2020-05-19 book/paper.tex
Fixed major bug in pointer handling
Fixed various typos

CXL is a concurrent programming language.  It was designed to teach
the basics of concurrent programming, but it is also useful for
testing new concurrent algorithms or even sequential and distributed
algorithms.  CXL programs are not intended to be ``run'' like programs
in most other programming languages---instead CXL programs are
``model checked'' to verify that the program has certain desirable
properties and does not suffer from bugs.  Model checking is
limited---it can only check programs that have ``finite state
spaces.''  But even programs that have infinite state spaces can
often be constrained (for example, by specifying a finite set of
possible inputs) so that model checking is possible.  If, for
example, a sorting algorithm is shown to work for any subset of
some set of say, 10, different values, then one can hope that it
will also work for sets that are larger than 10 values or sets that
use more than 10 values.

The syntax and semantics of CXL is similar to that of Python.
Python is familiar to many programmers and is easy to learn and
use.  We will assume that the reader is familiar with the basics
of Python programming.  We also will assume that the reader
understand some basics of machine architecture and how programs
are executed.

Like Python, CXL is an imperative,
dynamically typed and garbage collected programming language.
There are also some important differences:
\begin{itemize}
\item Every statement in CXL must be terminated by a semicolon
(even \texttt{while} statements and \texttt{def} statements);
in Python semicolons are optional and rarely used;
\item Correct indentation in CXL is encouraged but optional;
\item Python is object-oriented; CXL is not;
\item CXL does not support floating point;
\item CXL does not support I/O of any kind;
\item CXL does not support operator precedence or associativity.  Use
parentheses liberally to remove ambiguity.
\end{itemize}
There are also many unimportant ones that you will discover as
you get more familiar with programming in CXL.

@@

Figure~\ref{fig:triangle} gives a simple example of a CXL program.
It is sequential and has a method \texttt{square} that takes
an integer number as argument.  Each method has a variable called
\texttt{result} that eventually contains the result of the
method (there is no \texttt{return} statement in CXL).  The method
also has a variable called \texttt{n} containing the value of the
argument.  The \texttt{x..y} notation generates a set containing the numbers
from~\texttt{x} to~\texttt{y} (inclusive).  The last two lines in the program are
the most interesting.
They assign to \texttt{x} some unspecified value in the range \texttt{0..N},
and then verifies the \texttt{assert} statement that $\mathtt{triangle}(x)$
equals $x(x+1)/2$.

@@

\begin{quote}
Try it out!
\begin{itemize}
\item See what happens if, instead of initializing \texttt{result} to 0,
you initialize it to 1.
\item Write a similar program that computes squares.
\end{itemize}
\end{quote}

@@

Concurrent programming, aka multithreaded programming, involves multiple
processes running in parallel while sharing variables.
Figure~\ref{fig:inc} presents a simple example.  The program defines a
method \texttt{process} that takes a parameter called \texttt{self}.
It also initializes two shared variables \texttt{count} and \texttt{done}.
The \texttt{process()} method increments \texttt{count} and sets \texttt{done[self]}
to \texttt{True}.
Next the program spawns two processes, each invoking \texttt{process}.
Afterward there are three processes---after
all, there was already one, and it spawned two new ones.  We will call
the one that was already there the \emph{main process}.

@@

\begin{quote}
What will happen?
\begin{itemize}
\item Before you run the program, what do you think will happen?  Is the
program correct in that \texttt{count} will end up being 2?
\item What are the possible values of \texttt{count} at the time of the
\texttt{assert} statement?
(You may assume that \texttt{load} and \texttt{store} instructions of the
underlying architecture are atomic---in fact they are.)
\end{itemize}
\end{quote}

What is going on is that the CXL program is compiled to machine instructions,
and it is the machine instructions that are executed by the underlying
machine.  The details of this appear in the chapter~\ref{ch:cxlmachine},
but suffice it to
say that the machine has instructions that load values from memory and store
values into memory.  Importantly, it does not have instructions to atomically
increment or decrement values in memory locations.
So to increment a value in memory,
the machine must do at least three machine instructions.  Conceptually:
\begin{enumerate}
\item load the value from the memory location;
\item add 1 to the value;
\item store the value to the memory location.
\end{enumerate}

When running multiple processes, each essentially runs an instantiation of
the machine, and they do so in parallel.  As they execute, their machine
instructions are interleaved in unspecified and often random ways.
In fact, CXL will try \emph{all possible interleavings} of the processes
executing machine instructions.
The following is a possible interleaving of processes~0 and~1:
\begin{enumerate}
\item Process 0 loads the value of \texttt{count}, which is 0;
\item Process 1 loads the value of \texttt{count}, which is still 0;
\item Process 1 adds one to the value that it loaded (0), and
stores $1$ into \texttt{count};
\item Process 0 adds one to the value that it loaded (0), and
stores $1$ into \texttt{count};
\item Process 0 sets \texttt{upDone} to \texttt{True};
\item Process 1 sets \texttt{upDone} to \texttt{True}.
\end{enumerate}

@@

\begin{code}
\begin{verbatim}
    assert (count == 0) or (count == 1);
\end{verbatim}
\end{code}

@@

CXL makes extensive use of dictionaries.  Their syntax and properties
are a little different from Python.
A dictionary maps keys to values.
Unlike Python, any CXL value can be a key, including another
dictionary.
If \texttt{d} is a dictionary, and \texttt{k} is a key, then the
following expression retrieves the CXL value that $k$ maps to:
\begin{code}
\begin{verbatim}
d k
\end{verbatim}
\end{code}
This is unfamiliar, but in CXL square brackets can be used in the
same way as parentheses, so you can express it in the more familiar
form:
\begin{code}
\begin{verbatim}
d[k]
\end{verbatim}
\end{code}
However, if the key is an atom, say \texttt{.count}, then you
can write \texttt{d.count} instead of having to write
\texttt{d[.count]} (both will work).

Tuples are special forms of dictionaries where the keys are
the indexes into the tuple.  For example, the tuple
\texttt{(5, False)} is the same CXL value as
\texttt{dict\{ 0: 5, 1: False \}}.
The empty tuple \texttt{()} is the same value as \texttt{dict\{\}}.
As in Python, you can create singleton tuples by adding an
extra ",". For example, \texttt{(1,)}.

Again, square brackets and parentheses work the same in CXL, so
\texttt{[a, b, c]} (which looks like a Python list)
is the same as \texttt{(a, b, c)} (which looks like a Python tuple),
which in turn is the same as \texttt{dict\{ 0:a, 1:b, 2:c \}}.
So if \texttt{x == [ False, True ]},
then \texttt{x[0] == False} and \texttt{x[1] == True}.

Dictionaries have two handy unary operators.  If \texttt{d} is a
dictionary, then \texttt{keys d} (or equivalently \texttt{keys(d)})
returns the set of keys and \texttt{len d} returns the size of
this set.

@@

Before we delve into how to solve synchronization problems, it is important
to know a bit about the underlying machine.  A CXL program is translated
into a list of machine instructions that the machine executes.
The CXL machine is not an ordinary virtual machine, but its architecture
is nonetheless representative of conventional machines such as hardware
with memory and CPUs and virtual machines such as the Java Virtual Machine.

Instead of bits and bytes, a CXL machine manipulates CXL values.
A CXL machine has the following components:
\begin{itemize}
\item Code:  This is an immutable and finite list of CXL machine instructions,
generated from a CXL program.  The types of instructions will be described later.
\item Shared memory: A CXL machine has just one memory location containing
a CXL.
\item Processes:  Initially there is just the main process, but any process
can spawn an unbounded number of other processes and processes may terminate.
Each process has an immutable (but not necessarily unique) \emph{name tag},
a program counter, a stack of CXL values,
and a single mutable general purpose \emph{register} that contains a CXL.
\end{itemize}

@@

It may seem strange that there is only one memory location and that each
process has only one register.  However, this is not a limitation because
CXL values are unbounded trees.
By convention, both the memory and the register of a process always contain
a special type of CXL value in which the root is a dictionary
that maps strings to CXL values.  We call this a \emph{directory}.
A directory represents the state of a collection of variables named by the strings.

Directories can be organized into a tree.  Each node in the tree is then identified
by a sequence of strings, like a path name in the file system hierarchy.  We call
such a sequence the \emph{address} of a CXL value.

@@

The \texttt{Frame} instruction has three arguments:
\begin{enumerate}
\item A token containing the name of the method;
\item A token containing the name of the argument of the method;
\item A program counter of the method's \texttt{Return} instruction.
\end{enumerate}

A \emph{token} is a four-tuple containing a \emph{lexeme}, a file name, a
line number in the file, and a column number in the file.  For example,
\texttt{(’up’, ’Up.cxl’, 1, 5)} represents the lexeme \texttt{up} in
file \texttt{Up.xcl}, line 1, column 5.  Tokens make it easy to relate
the CXL machine instructions to the CXL source code.

@@

Figure~\ref{fig:incoutput} shows the output produced by running the
\texttt{Up.cxl} program.
It starts by reporting that the assertion on line 12 failed and that
\texttt{count} has value 1 instead of 0:
\texttt{Assertion failed (’assert’, ’Up.cxl’, 12, 1) 1}.
Next it reports an execution that failed this assertion.  The output has
five columns:
\begin{enumerate}
\item The entry point of the process (\texttt{\_\_main\_\_} for the main process);
\item The tag of the process (in this case always 0);
\item The sequence of program counters of the CXL machine instructions that the process executed;
\item The current program counter of the process;
\item The contents of the shared memory.
\end{enumerate}

@@

Unfortunately, if we run the program, we find that the assertion still
fails.  Diagnosing the problem, we see that the \texttt{lock} variable
suffers from the same problem as the \texttt{count} variable
in Figure~\ref{fig:inc}: operations
on it consist of several instructions are divisible.  It is thus possible
for both processes to believe the lock is available and obtain the lock
at the same time.

Figure~\ref{fig:upflags} presents a solution based on each process having
a flag indicating that it is trying to enter the critical section.
A process can write its own flag and read the flag of its peer.
After setting its flag, the process waits until the other process
(\texttt{1 - self}) is not trying to enter the critical section.
If we run this program, the assertion does not fail.  In fact, this
solution does prevent both processes being in the critical section at
the same time.
However, it has a different problem called livelock: if both try to
enter the critical section at the same time, they may end up waiting for
one another.

@@

A method \texttt{m} with argument \texttt{a} is invoked in its
most basic form as follows (assigning the result to \texttt{x}).
\begin{code}
\begin{verbatim}
var x = m a;
\end{verbatim}
\end{code}
That's right, no parentheses are required.  In fact, if you invoke
\texttt{m(a)}, the argument is \texttt{(a)}, which is the same
as \texttt{a}.
If you invoke \texttt{m()}, the argument is \texttt{()},
which is the empty tuple.  If you invoke
If you invoke \texttt{m(a, b)}, the argument is \texttt{(a, b)},
the tuple consisting of values \texttt{a} and \texttt{b}.
You have already seen examples of this: the expression
\texttt{atLabel.cs} invokes method \texttt{atLabel} with
argument \texttt{.cs}.

>>>>

CXL is a concurrent programming language.  It was designed to teach
the basics of concurrent programming, but it is also useful for
testing new concurrent algorithms or even sequential and distributed
algorithms.  CXL programs are not intended to be ``run'' like programs
in most other programming languages---instead CXL programs are
``model checked'' to verify that the program has certain desirable
properties and does not suffer from bugs.  Model checking is
limited---it can only check programs that have ``finite state
spaces.''  But even programs that have infinite state spaces can
often be constrained (for example, by specifying a finite set of
test inputs) so that model checking is possible.  If, for
example, a sorting algorithm is shown to work for any subset of
some set of say, 5, different values, then one can hope that it
will also work for sets that are larger than 5 values or sets that
use more than 5 different values.

The syntax and semantics of CXL is similar to that of Python.
Python is familiar to many programmers and is easy to learn and
use.  We will assume that the reader is familiar with the basics
of Python programming.  We also will assume that the reader
understand some basics of machine architecture and how programs
are executed.  For example, we assume that the reader is familiar
with the concepts of CPU, memory, register, and stack.

Like Python, CXL is an imperative,
dynamically typed and garbage collected programming language.
There are also some important differences:
\begin{itemize}
\item Every statement in CXL must be terminated by a semicolon
(even \texttt{while} statements and \texttt{def} statements);
in Python semicolons are optional and rarely used;
\item Correct indentation in CXL is encouraged but optional;
\item Python is object-oriented; CXL is not;
\item CXL does not (currently) support floating point;
\item CXL does not support I/O of any kind;
\item CXL does not support operator precedence or associativity.  Use
parentheses liberally to remove ambiguity.
\end{itemize}
There are also many unimportant ones that you will discover as
you get more familiar with programming in CXL.

@@

Figure~\ref{fig:triangle} gives a simple example of a CXL program.
It is sequential and has a method \texttt{square} that takes
an integer number as argument.  Each method has a variable called
\texttt{result} that eventually contains the result of the
method (there is no \texttt{return} statement in CXL).  The method
also has a variable called \texttt{n} containing the value of the
argument.  The \texttt{x..y} notation generates a set containing the numbers
from~\texttt{x} to~\texttt{y} (inclusive).  The last two lines in the program are
the most interesting.
They assign to \texttt{x} some unspecified value in the range \texttt{0..N}
and then verifies that $\mathtt{triangle}(x)$ equals $x(x+1)/2$.

@@

\begin{quote}
Try it out!
\begin{itemize}
\item See what happens if, instead of initializing \texttt{result} to 0,
you initialize it to 1.
\item Write a similar program that computes squares or factorials.
\end{itemize}
\end{quote}

@@

Concurrent programming, aka multithreaded programming, involves multiple
processes running in parallel while sharing variables.
Figure~\ref{fig:inc} presents a simple example.  The program defines a
method \texttt{process} that takes a parameter called \texttt{self}.
It also initializes two shared variables \texttt{count} and \texttt{done}.
The \texttt{process(self)} method increments \texttt{count} and sets \texttt{done[self]}
to \texttt{True}.
Next the program spawns two processes, each invoking \texttt{process}.
Afterward there are three processes---after
all, there was already one, and it spawned two new ones.  We will call
the one that was already there the \emph{main process}.

@@

\begin{quote}
What will happen?
\begin{itemize}
\item Before you run the program, what do you think will happen?  Is the
program correct in that \texttt{count} will always end up being 2?
\item What are the possible values of \texttt{count} at the time of the
\texttt{assert} statement?
(You may assume that \texttt{load} and \texttt{store} instructions of the
underlying architecture are atomic---in fact they are.)
\end{itemize}
\end{quote}

What is going on is that the CXL program is compiled to machine instructions,
and it is the machine instructions that are executed by the underlying CXL
machine.  The details of this appear in Chapter~\ref{ch:cxlmachine},
but suffice it to
say that the machine has instructions that load values from memory and store
values into memory.  Importantly, it does not have instructions to atomically
increment or decrement values in memory locations.
So to increment a value in memory,
the machine must do at least three machine instructions.  Conceptually:
\begin{enumerate}
\item load the value from the memory location;
\item add 1 to the value;
\item store the value to the memory location.
\end{enumerate}

When running multiple processes, each essentially runs an instantiation of
the machine, and they do so in parallel.  As they execute, their machine
instructions are interleaved in unspecified and often random ways.
A program is correct if it works for any interleaving.
In fact, CXL will try all possible interleavings of the processes
executing machine instructions.

The following is a possible interleaving of processes~0 and~1:
\begin{enumerate}
\item Process 0 loads the value of \texttt{count}, which is 0;
\item Process 1 loads the value of \texttt{count}, which is still 0;
\item Process 1 adds one to the value that it loaded (0), and
stores $1$ into \texttt{count};
\item Process 0 adds one to the value that it loaded (0), and
stores $1$ into \texttt{count};
\item Process 0 sets \texttt{upDone} to \texttt{True};
\item Process 1 sets \texttt{upDone} to \texttt{True}.
\end{enumerate}

@@

\begin{code}
\begin{verbatim}
    assert (count == 1) or (count == 2);
\end{verbatim}
\end{code}

@@

CXL makes extensive use of dictionaries.  Their syntax and properties
are a little different from Python.
A dictionary maps keys to values.
Unlike Python, any CXL value can be a key, including another
dictionary.
If \texttt{d} is a dictionary, and \texttt{k} is a key, then the
following expression retrieves the CXL value that $k$ maps to:
\begin{code}
\begin{verbatim}
d k
\end{verbatim}
\end{code}
This is unfamiliar, but in CXL square brackets can be used in the
same way as parentheses, so you can express it in the more familiar
form:
\begin{code}
\begin{verbatim}
d[k]
\end{verbatim}
\end{code}
However, if \texttt{d = dict\{ .count: 3 \}}, then you can write
can write \texttt{d.count} (which has value 3) instead of having to write
\texttt{d[.count]} (although both will work).
The meaning of \texttt{d a b c ...} is \texttt{((((d a) b) c) ...)}.

Tuples are special forms of dictionaries where the keys are
the indexes into the tuple.  For example, the tuple
\texttt{(5, False)} is the same CXL value as
\texttt{dict\{ 0: 5, 1: False \}}.
The empty tuple \texttt{()} is the same value as \texttt{dict\{\}}.
Note that this is different from the empty set, which is \texttt{\{\}}.
As in Python, you can create singleton tuples by adding an
extra ",". For example, \texttt{(1,)}.

Again, square brackets and parentheses work the same in CXL, so
\texttt{[a, b, c]} (which looks like a Python list)
is the same CXL value as \texttt{(a, b, c)} (which looks like a Python tuple),
which in turn is the same CXL value as \texttt{dict\{ 0:a, 1:b, 2:c \}}.
So if \texttt{x == [ False, True ]},
then \texttt{x[0] == False} and \texttt{x[1] == True}, just like in Python.

CXL is not an object-oriented language, however, so objects don't have
built-in methods.  However, CXL does have some powerful operators to
make up for some of that.
For example, dictionaries have two handy unary operators.
If \texttt{d} is a
dictionary, then \texttt{keys~d} (or equivalently \texttt{keys(d)})
returns the set of keys and \texttt{len~d} returns the size of
this set.

@@

Before we delve into how to solve synchronization problems, it is important
to know a bit about the underlying machine.  A CXL program is translated
into a list of machine instructions that the CXL machine executes.
The CXL machine is not an ordinary virtual machine, but its architecture
is nonetheless representative of conventional machines such as hardware
with memory and CPUs and virtual machines such as the Java Virtual Machine.

Instead of bits and bytes, a CXL machine manipulates CXL values.
A CXL machine has the following components:
\begin{itemize}
\item Code:  This is an immutable and finite list of CXL machine instructions,
generated from a CXL program.  The types of instructions will be described later.
\item Shared memory: A CXL machine has just one memory location containing
an CXL value.
\item Processes:  Initially there is just the main process, but any process
can spawn an unbounded number of other processes and processes may terminate.
Each process has an immutable (but not necessarily unique) \emph{name tag},
a program counter, a stack of CXL values,
and a single mutable general purpose \emph{register} that contains a CXL value.
\end{itemize}

@@

It may seem strange that there is only one memory location and that each
process has only one register.  However, this is not a limitation because
CXL values are unbounded trees.
Both the memory and the register of a process always contain
a special type of CXL value in which the root is a dictionary
that maps atoms to CXL values.  We call this a \emph{directory}.
A directory represents the state of a collection of variables named by the atoms.

Because directories are CXL values themselves,
directories can be organized into a tree.
Each node in the tree is then identified
by a sequence of atoms, like a path name in the file system hierarchy.  We call
such a sequence the \emph{address} of a CXL value.

@@

The \texttt{Frame} instruction has three arguments:
\begin{enumerate}
\item A token containing the name of the method;
\item A list of tokens containing the names of the arguments of the method;
\item A program counter of the method's \texttt{Return} instruction.
\end{enumerate}

A \emph{token} is a four-tuple containing a \emph{lexeme}, a file name, a
line number in the file, and a column number in the file.  For example,
\texttt{(’up’, ’Up.cxl’, 1, 5)} represents the lexeme \texttt{up} in
file \texttt{Up.xcl}, line~1, column~5.  Tokens make it easy to relate
the CXL machine instructions to the CXL source code.

@@

Figure~\ref{fig:incoutput} shows the output produced by running the
\texttt{Up.cxl} program.
It starts by reporting that the assertion on line 12 failed and that
\texttt{count} has value 1 instead of 0:
\texttt{Assertion failed (’assert’, ’Up.cxl’, 12, 1) 1}.
Next it reports an execution that failed this assertion.  The output has
four columns:
\begin{enumerate}
\item The name tag of the process;
\item The sequence of program counters of the CXL machine instructions that the process executed;
\item The current program counter of the process;
\item The contents of the shared memory.
\end{enumerate}

@@

Unfortunately, if we run the program, we find that the assertion still
fails.  Diagnosing the problem, we see that the \texttt{lock} variable
suffers from the same problem as the \texttt{count} variable
in Figure~\ref{fig:inc}: operations
on it consist of several instructions.  It is thus possible
for both processes to believe the lock is available and to obtain the lock
at the same time.

Figure~\ref{fig:upflags} presents a solution based on each process having
a flag indicating that it is trying to enter the critical section.
A process can write its own flag and read the flag of its peer.
After setting its flag, the process waits until the other process
(\texttt{1 - self}) is not trying to enter the critical section.
If we run this program, the assertion does not fail.  In fact, this
solution does prevent both processes being in the critical section at
the same time.
However, it has a different problem called \emph{livelock}: if both try to
enter the critical section at the same time, they may end up waiting for
one another.

@@

A method \texttt{m} with argument \texttt{a} is invoked in its
most basic form as follows (assigning the result to \texttt{x}).
\begin{code}
\begin{verbatim}
var x = m a;
\end{verbatim}
\end{code}
That's right, no parentheses are required.  In fact, if you invoke
\texttt{m(a)}, the argument is \texttt{(a)}, which is the same
as \texttt{a}.
If you invoke \texttt{m()}, the argument is \texttt{()},
which is the empty tuple.
If you invoke \texttt{m(a, b)}, the argument is \texttt{(a, b)},
the tuple consisting of values \texttt{a} and \texttt{b}.
You have already seen examples of this: the expression
\texttt{atLabel.cs} invokes method \texttt{atLabel} with
argument \texttt{.cs}.
